Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/splitting.tex

\paragraph {Unit.} 
As a simple first case, consider splitting the unit value, $\tup{}$
which masks to $\mval {\tup{}} {\tup{}}$.
%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
Splitting must produce a $c$ and $l.r$ satisfying,
\[
	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
Splitting $\pure{e}$ follows in similar fashion, except the first 
component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.

\paragraph {Injections.}
Now consider splitting the non-terminal, $\inl {e}$, where $\diaone
{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$. Therefore:
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
splitting must produce a $c'$ and $l.r'$ satisfying,
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
\]
Our algorithm generates a first pass $c'$ that produces the same
boundary value as splitting $e$:
\[
c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
\]
Splitting the injection also generates the same resumer as splitting
the subterm $e$: ($l.r' = l.r$).

\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
where $\diaone {e_k} {\xi_k} {v_k}$ and 
$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
Inductively:
\[
	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
\]
Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
splitting generates:
\[
	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
\]
The resulting boundary value is the tuple of the boundary values of
the subterms, $b = \tup{b_1,b_2}$.  The value is 
constructed using,
\[
c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
\]
And deconstructed in the resumer with the tuple pattern,
\[
l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
\]
\paragraph {Case.}
Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
Here we consider only the case where
$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
(The \texttt{inr} case is analogous.)

\noindent
Inductively,
\[
	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
\]
Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
Therefore:
\[
	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2} {\tup{i_2,b_2}} 
\]
\[
\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
\]
(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
of $v_1$ into $c_2$ and $l_2.r_2$.)

@@

\paragraph {Function.} 
Function introduction has a $\tup{}$ boundary value, since functions
are already fully reduced in our semantics.  However, since the body
of a function may itself be multistage, splitting must continue into
it.  The first pass generated by split produces an $i$ that is a new
function formed from the first-stage part of the original body.  The
resumer is a new function formed out of the second-stage part of the
original body.  It is the responsibility of the application site to
save the precomputation of the function body and pass it to the
resumer version of the function.

@@

Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
  \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
Note that for any $\diatwo e q$, the residual $q$ must have the form
$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even under binders and
for nullary terms.  For each $k \in \{1,\ldots,n\}$ splitting the
subterm $e_k$ yields $p_k$ and $l_k.r_k$ where:
\[
	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
Correspondingly, splitting the $n$-ary $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where:
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

Again, this holds even for nullary terms and under binders.
In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
\[
\pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
\]
Note how the precomputations of $e_2$ and $e_3$ are pulled out from
underneath the binder into a tuple that is always executed in the
first stage. This occurs regardless of what branch is eventually
executed in the second stage.

\paragraph{Next and Prev.}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
their splitting rules are relatively simple.
Given $\diatwo e q$, splitting yields $p$ and $l.r$ where:
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where:
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.

>>>>

\paragraph {Unit.} 
As a simple first case, consider splitting the unit value, $\tup{}$
which masks to $\mval {\tup{}} {\tup{}}$.
%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
Splitting must produce a $c$ and $l.r$ satisfying
\[
	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
Splitting $\pure{e}$ follows in similar fashion, except the first 
component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.

\paragraph {Injections.}
Now consider splitting the non-terminal $\inl {e}$, where $\diaone
{e} \xi v$ and $\rtab \xi v \vsplito \mval i q$, and thus:
\[
	\splitone e A c {l.r} \text{ and } \reduce c {\tup{i,b}} \text{ and } [b/l]r \equiv q
\]
Since $\diaone {\inl e} \xi {\inl v}$ and $\rtab \xi {\inl v} \vsplito \mval {\inl i} q$,
splitting must produce a $c'$ and $l.r'$ satisfying
\[
	\reduce {c'} {\tup{\inl i,b'}} \text{ and } [b'/l]r' \equiv q
\]
Our algorithm generates a first pass $c'$ that produces the same
boundary value as splitting $e$:
\[
c' = \letin {\tup {x,y}} c {\tup{\inl x, y}}
\]
Splitting the injection also generates the same resumer as splitting
the subterm $e$ (that is, $l.r' = l.r$).

\paragraph{Tuple.} For an example of splitting terms with more than one argument, consider the tuple $\tup{e_1,e_2}$,
where $\diaone {e_k} {\xi_k} {v_k}$ and 
$\rtab {\xi_k} {v_k} \vsplito \mval {i_k} {q_k}$ for $k = 1,2$.  
Inductively:
\[
	\splitone {e_k} A {c_k} {l_k.r_k} \text{ and } \reduce {c_k} {\tup{i_k,b_k}} \text{ and } [b_k/l_k]r_k \equiv q_k
\]
Since $\diaone {\tup{e_i,e_2}} {\xi_1,\xi_2} {\tup{v_1,v_2}}$ 
and $\rtab {\xi_1,\xi_2} {\tup{v_1,v_2}}$ and $\tup{v_1,v_2} \vsplito \mval {\tup{i_1,i_2}} {\tup{q_1,q_2}}$,
splitting generates:
\[
	\reduce {c} {\tup{\tup{i_1,i_2},b}} \text{ and } [b/l]r \equiv \tup{q_1,q_2}
\]
The resulting boundary value is the tuple of the boundary values of
the subterms, $b = \tup{b_1,b_2}$.  The value is 
constructed using
\[
c = \letin{\tup{y_1,z_1}}{c_1}{\letin{\tup{y_2,z_2}}{c_2}{\tup{\tup{y_1, y_2},\tup{z_1, z_2}}}}
\]
and deconstructed in the resumer with a tuple pattern:
\[
l.r = \tup{l_1,l_2}.\tup{r_1,r_2}
\]
\paragraph {Case.}
Splitting is most complicated at $\caseof {e_1} {x_2.e_2} {x_3.e_3}$ terms because of their divergent control flow.
Here we consider only the case where
$\diaone {e_1} {\xi_1} {\inl {v_1}}$ and \\ $\rtab {\xi_1} {\inl {v_1}} \vsplito \mval {\inl {i_1}} {q_1}$.
(The \texttt{inr} case is analogous.)

\noindent
Inductively,
\[
	\splitone {e_1} A {c_1} {l_1.r_1} \text{ and } \reduce {c_1} {\tup{\inl {i_1},b_1}} \text{ and } [b_1/l_1]r_1 \equiv q_1
\]
Additionally, the first branch evaluates $\diaone {[v_1/x_2]e_2} {\xi_2} {v_2}$
and $\rtab {\xi_1,\xi_2} {v_2} \vsplito \mval {i_2} {q_2}$.
Therefore:
\begin{gather*}
	\splitone {e_2} A {c_2} {l_2.r_2} \text{ and } \reduce {[i_1/x_2]c_2}
  {\tup{i_2,b_2}} \\
\text{ and } [q_1/x_2][b_2/l_2]r_2 \equiv q_2
\end{gather*}
(Because $e_2$ is open on $x_2$, $\pipeM{c_2}{l_2}{r_2}$ is also open on $x_2$.
When we substitute some $v_1$ for $x_2$ in $e_2$, we must substitute the masks
of $v_1$ into $c_2$ and $l_2.r_2$.)

@@

\paragraph {Function.} 
Function introduction has a $\tup{}$ boundary value, since functions
are already fully reduced in our semantics.  However, since the body
of a function may itself be multistage, splitting must descend into
it.  The first pass generated by splitting produces an $i$ that is a new
function formed from the first-stage part of the original body.  The
resumer is a new function formed out of the second-stage part of the
original body.  It is the responsibility of the application site to
save the precomputation of the function body and pass it to the
resumer version of the function.

@@

Consider any $n$-ary \bbtwo-term $e = \scriptCapp {e_1 \ttsemi \ldots
  \ttsemi e_n}$ (except \texttt{prev}, which is discussed shortly).
Note that for any $\diatwo e q$, the residual $q$ must have the form
$\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$, even if $\mathcal{C}$ is
nullary or the $q_i$ have binders.
For each $k$, splitting the subterm $e_k$ yields $p_k$ and $l_k.r_k$ where
\[
	\reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
Correspondingly, splitting $\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$ generates $p$ and $l.r$ where
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

%Again, this holds even for nullary terms and under binders.
In particular, $\caseof{e_1}{x_2.e_2}{x_3.e_3}$ splits into:
\[
\pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}
\]
Note how the precomputations of $e_2$ and $e_3$ are pulled out from
underneath the binder into a tuple that is always executed in the
first stage. This occurs regardless of what branch is eventually
executed in the second stage.

\paragraph{Staging constructs.}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second-stage terms at \bbtwo\ and encapsulated second-stage terms at \bbonem,
their splitting rules are relatively simple.
Given $\diatwo e q$, splitting yields $p$ and $l.r$ where
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
Then for $\diaone {\next {e}} {{\hat y} \mapsto q} {\next {\hat y}}$
and\\ $\rtab {{\hat y} \mapsto q} {\next {\hat y}} \vsplito \mval {\tup{}} {\letin {\hat y} q {\hat y}}$, splitting must produce $c'$ and $l'.r'$ where
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin {\hat y} q {\hat y}
\]
Our algorithm chooses $c' = \tup{\tup{},p}$ and $l'.r' = l.r$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
